---
title: "Alpha–beta–gamma diversity over time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alpha--beta--gamma diversity over time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
```

```{r setup}
library(partyscape)
```

Multiplicative beta diversity (Jost 2007) answers: "how many distinct
compositions is this trajectory equivalent to?" A country whose seat
shares drift gently has beta near 1; a country with genuinely different
elections has beta approaching T.

## Stable country

```{r}
stable <- rbind(c(0.45, 0.30, 0.15, 0.10),
                c(0.44, 0.31, 0.14, 0.11),
                c(0.46, 0.29, 0.16, 0.09))
alpha_beta_gamma(stable, q = c(0, 1, 2, 5))
```

## Churning country

```{r}
churn <- rbind(c(0.60, 0.20, 0.15, 0.05),
               c(0.30, 0.30, 0.25, 0.15),
               c(0.15, 0.25, 0.35, 0.25))
bd <- alpha_beta_gamma(churn, q = c(0, 1, 2, 5))
bd
plot(bd)
```

## Collapsing to elections

When consecutive country-years repeat the same composition between
elections, pass `years` to weight by term length:

```{r}
history <- rbind(c(0.45, 0.30, 0.15, 0.10),   # 2000
                 c(0.45, 0.30, 0.15, 0.10),   # 2001 — same
                 c(0.40, 0.30, 0.20, 0.10))   # 2002 — new
alpha_beta_gamma(history, q = 2, years = c(2000, 2001, 2002))
```
