---
title: "Profile crossings: when ENP isn't enough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profile crossings: when ENP isn't enough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
```

```{r setup}
library(partyscape)
```

Two party systems can share an effective-number-of-parties value and yet
look nothing alike at the ends of the profile. `crossings()` detects the
q values where one system overtakes the other.

The example below uses the paper's canonical case — Netherlands 1982
vs. Sweden 2002. Both have ENP near 4, but the Dutch system has many
small parties (steep profile) while Sweden has fewer, larger ones (flat
profile), so the profiles cross twice.

```{r canonical-crossing}
# Party-labeled vectors. Column order is party identity (stable, e.g.,
# alphabetical within each system), NOT rank by size.
netherlands_1982 <- c(
  PvdA = 0.313, CDA = 0.300, VVD = 0.240, D66 = 0.040,
  PSP = 0.020, CPN = 0.020, SGP = 0.020, PPR = 0.013,
  RPF = 0.013, GPV = 0.007, CP  = 0.007, DS70 = 0.007
)
sweden_2002 <- c(
  SAP = 0.415, Moderata = 0.156, Folkpartiet = 0.138,
  Kristdemokraterna = 0.096, Vansterpartiet = 0.086,
  Centerpartiet = 0.063, Miljopartiet = 0.046
)

c(enp_nl = enp(netherlands_1982), enp_se = enp(sweden_2002))

cr <- crossings(netherlands_1982, sweden_2002, q = seq(0, 10, 0.01))
cr
plot(cr)
```

## Dataset-wide scans

For large comparisons, `method = "signchange"` is cheap and robust. The
rows below are draws from an asymmetric Dirichlet — a stand-in for real
party-labeled compositions; the point is that each row is a single
party system, and the `crossings()` call treats each row as its own
system without any sorting assumption.

```{r scan}
set.seed(1)
many <- t(replicate(8, { x <- rgamma(5, 0.8); x / sum(x) }))
colnames(many) <- paste0("Party_", LETTERS[1:5])
rownames(many) <- paste0("sys", 1:8)
cr_all <- crossings(many, q = seq(0, 5, 0.01), method = "signchange")
head(as.data.frame(cr_all))
```
