---
title: "End-to-end with ParlGov data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-end with ParlGov data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_net <- requireNamespace("curl", quietly = TRUE) &&
  isTRUE(tryCatch(curl::has_internet(), error = function(e) FALSE))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4,
                      eval = has_net)
```

```{r setup}
library(partyscape)
```

This vignette walks through the full pipeline on live
[ParlGov](https://www.parlgov.org/) data. On first run it downloads the
current `view_election` CSV and caches it in
`tools::R_user_dir("partyscape", "cache")`.

```{r fetch}
pg <- fetch_parlgov()
head(pg[, c("country", "date", "party", "seats")])
```

## Reshape to a seat-share matrix

```{r reshape}
shares <- parlgov_seat_shares(pg,
                              countries = c("Netherlands", "Sweden"),
                              years     = 1960:2020,
                              type      = "parliament")
dim(shares)
```

## Pillar 1 — diversity profiles for a handful of elections

```{r profile}
pick <- c("Netherlands_1982", "Sweden_2002")
plot(diversity_profile(shares[rownames(shares) %in% pick, , drop = FALSE]))
```

## Pillar 2 — crossings between the two pick elections

```{r crossings}
cr <- crossings(shares["Netherlands_1982", ],
                shares["Sweden_2002", ],
                q = seq(0, 5, 0.01))
cr
plot(cr)
```

## Pillar 3 — beta diversity of one country's trajectory

```{r beta}
nl <- shares[attr(shares, "country") == "Netherlands", , drop = FALSE]
yr <- attr(shares, "year")[attr(shares, "country") == "Netherlands"]
plot(alpha_beta_gamma(nl, q = c(0, 1, 2, 5), years = yr))
```

## Citation

If you use ParlGov data, please cite Döring & Manow (2024), Harvard
Dataverse, DOI
[10.7910/DVN/2VZ5ZC](https://doi.org/10.7910/DVN/2VZ5ZC).
