---
title: "Getting started with partyscape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with partyscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
```

```{r setup}
library(partyscape)
```

## Party-labeled compositions

`partyscape` works on **party-labeled** compositions: each column is a
specific party, preserved across elections. The package deliberately
does *not* ask you to row-sort shares. Sorting by size makes "column 1"
mean "the largest party that year," which loses identity information
when parties swap positions across elections — the behaviour the
package's `alpha_beta_gamma()` is designed to detect.

Concretely: use named vectors (or a matrix with named columns). The
entry order reflects party identity, not rank.

```{r single}
germany_2009 <- c(CDU = 0.312, CSU = 0.072, SPD = 0.235,
                  FDP = 0.150, Gruene = 0.109, Linke = 0.122)

hill_number(germany_2009, c(0, 1, 2, Inf))
enp(germany_2009)
dominance_gap(germany_2009)
top_two(germany_2009)
```

`diversity_profile()` sweeps q over a grid and returns an object with
`print()` / `plot()` / `autoplot()` methods:

```{r profile-plot}
dp <- diversity_profile(germany_2009)
plot(dp)
```

## Several systems at once

Rows are systems, columns are parties. Column order is stable across
rows — party identity is preserved. Here are three German Bundestag
elections. Notice SPD > CDU/CSU in 2002 and CDU/CSU > SPD in 2005/2009:
the columns are a fixed set of parties, NOT ordered by size.

```{r multi}
germany <- rbind(
  `2002` = c(CDU = 0.315, CSU = 0.096, SPD = 0.416,
             FDP = 0.078, Gruene = 0.091, Linke = 0.003),
  `2005` = c(CDU = 0.293, CSU = 0.075, SPD = 0.362,
             FDP = 0.099, Gruene = 0.083, Linke = 0.088),
  `2009` = c(CDU = 0.312, CSU = 0.072, SPD = 0.235,
             FDP = 0.150, Gruene = 0.109, Linke = 0.122)
)
plot(diversity_profile(germany))
```

Evenness profile — scaled so every curve starts at 1 at q = 0:

```{r evenness}
plot(evenness_profile(germany))
```

## Two different-party systems

When two systems don't share a party set (e.g., Germany and the UK),
compute each profile separately and plot them alongside each other —
padding different-length vectors with zeros would distort neither
profile but would invite the row-sorted mental model that the package
tries to avoid.

```{r different-parties}
uk_1997 <- c(Labour = 0.634, Conservative = 0.250, LibDem = 0.070,
             UUP = 0.015, SNP = 0.009, PlaidCymru = 0.006,
             SDLP = 0.005, DUP = 0.003, SinnFein = 0.003,
             Independent = 0.005)
plot(diversity_profile(uk_1997,      id = "UK 1997"))
plot(diversity_profile(germany_2009, id = "Germany 2009"))
```

## What to read next

- `vignette("crossings")` — when ENP isn't enough.
- `vignette("beta")` — alpha-beta-gamma over time.
- `vignette("parlgov")` — end-to-end with live ParlGov data.
- `vignette("replication")` — replicate every result from the paper.
