---
title: "Getting started with partyscape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with partyscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
```

```{r setup}
library(partyscape)
```

## One composition at a time

A composition is any numeric vector of non-negative shares summing
(roughly) to 1; raw seat counts are accepted — they are renormalized with
a warning.

```{r}
p <- c(0.45, 0.30, 0.15, 0.10)
hill_number(p, c(0, 1, 2, Inf))
enp(p); dominance_gap(p); top_two(p)
```

`diversity_profile()` sweeps q over a grid.

```{r}
dp <- diversity_profile(p)
plot(dp)
```

## Several compositions at once

Rows of a matrix are treated as separate systems:

```{r}
m <- rbind(a = c(0.4, 0.3, 0.2, 0.1),
           b = c(0.6, 0.2, 0.1, 0.1),
           c = c(0.3, 0.3, 0.2, 0.2))
plot(diversity_profile(m))
plot(evenness_profile(m))
```

## What to read next

- `vignette("crossings")` — when ENP isn't enough.
- `vignette("beta")` — alpha-beta-gamma over time.
- `vignette("parlgov")` — end-to-end with live ParlGov data.
